Open-Source machine learning BANTER acoustic classification of a cryptic echolocating species

### 2023-02-07 manuscript/manuscript.tex
Update manuscript and hawaii data
Extensive work on Manuscript and preparation of Hawaii data (in folder). Hawaii data *should* be good to knit. Manuscript has references, needs updated results, figures, inline data, and updated discussion.

\title{Open-Source machine learning BANTER acoustic classification of a
cryptic echolocating species}
\author{Shannon Rankin \and Taiki Sakai \and Frederick Archer \and Jay
Barlow \and Danielle Cholewiak \and Annamaria DeAngelis \and Jennifer L.
K. McCullough \and Erin Oleson \and Melissa Soldevilla}
\date{12/12/22}

\begin{document}
\maketitle
\begin{abstract}
Passive acoustic monitoring is increasingly used for assessing
populations of marine mammals; however, analysis of large datasets is
limited by our ability to easily classify sounds detected.
Classification of beaked whale acoustic events, in particular, require
evaluation of multiple lines of evidence by expert analysts. Here we
present a highly automated approach to acoustic detection and
classification using supervised machine learning and open source
software methods. Data from four large scale surveys of beaked whales
(North Atlantic, South Atlantic, Hawaii, and Eastern Pacific) were
analyzed PAMGuard (acoustic detection), PAMpal (acoustic analysis) and
BANTER (hierarchical random forest classifier). Overall correct
classification results ranged from 86\% for the South Atlantic data to
99\% for the North Atlantic. Results for many species could likely be
improved with increased sample sizes and consideration of alternative
automated detectors. These methods provide a highly automated approach
to acoustic detection and classification using open source methods that
can be readily adopted for species and geographic regions.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, borderline west={3pt}{0pt}{shadecolor}, interior hidden, breakable, enhanced, frame hidden]}{\end{tcolorbox}}\fi

@@

Passive acoustic monitoring (PAM) from has proven to be a valuable tool
for studying populations of marine mammals (Van Parijs et al., 2009);
however, the value of these studies depends on our ability to identify
the sources of the sounds we are monitoring. Historically, experienced
acousticians manually identify stereotyped sounds that could be reliably
attributed to a given species, based on their spectral or temporal
characteristics (Baumann-Pickering et al., 2013; Bittle and Duncan,
2013; Rankin and Barlow, 2005; Soldevilla et al., 2008). The dramatic
increase in recordings make it impossible for experienced acousticians
to manually annotate all data, and there has been an increasing need for
the development of automated classification routines that can provide
accurate determinations of the species responsible for the sounds.

Beaked whales are deep diving marine mammals found in offshore waters;
their long dive intervals and cryptic surfacing behavior make them
difficult to study using typical shipboard visual observation methods
(MacLeod, 2018). However, beaked whales make stereotyped echolocation
clicks that have temporal and spatial characteristics that can be used
to differentiate species (Baumann-Pickering et al.~2013). Detection of
beaked whale signals in large datasets is greatly expanding our
understanding of their population structure and the potential impact of
human activities on these species (Barlow et al., 2021a;
Baumann-Pickering et al., 2014; Simonis et al., 2020). Unfortunately,
manual classification of beaked whale echolocation clicks requires
analysis by trained observers using a number of visual aids to examine
the call characteristics. Development of automated classification
routines, if accurate, serve to improve the efficiency and decrease the
cost of analyzing large datasets.

BANTER is a supervised machine learning acoustic event classifier
originally developed for classifying dolphin schools from acoustic
recordings (Rankin et al., 2017). This hierarchical random forest event
classifier relies on an initial call classifier (in the dolphin case
study, this included one each for whistles, echolocation clicks, and
burst pulses) followed by an event classifier. The event classifier
considers the probability of assignment to species for each call in the
call classifier, along with any event-level characteristics (such as
call rate). BANTER is very flexible and can accommodate any number of
measures from any number of detectors. While BANTER relies on multiple
call types, relatively minor changes to a call detector could yield
different detector results (Rankin et al., 2017). In this case, settings
for a whistle \& moan detector were modified to improve performance at
detecting burst pulses. This approach of applying multiple call
detectors (of the same type but with different settings) suggests that
perhaps BANTER could be used on species that only produce one call type
(or where only one call type was analyzed), as long as the settings of
the automated call detector were modified such that the overall number
of detections was different. This approach was successfully applied to
classified echolocation clicks for narwhal and belugas (Zahn et al.,
2021). For species who primarily (or exclusively) produce echolocation
clicks, BANTER may serve as an option for automated machine learning
classification.

@@

Recordings from the western North Atlantic Ocean (U.S. east coast) were
collected using towed hydrophone arrays during the 2016 AMAPPS II
(Atlantic Marine Assessment Program for Protected Species) survey
(cite). AMAPPS II was subdivided into the northern (NAtlantic) and
southern (SAtlantic) survey area. The NAtlantic study area ranged from
Massachusetts south to New Jersey (HB1603, cite) and recordings used in
this analysis consisted of a single hydrophone (HTI-96-min, High Tech
Inc., Long Beach, MS) recorded at 192 kHz sample rate with a 1 kHz high
pass filter (National Instruments USB-6356 A/D card, see DeAngelis et
al., 2018 for more information). The SAtlantic area ranged from Delaware
south to central Florida (GU1605, cite) and recordings used in this
analysis consisted of a single hydrophone (Reson etc) recorded 500 kHz
sample rate (soundcard) and decimated to 192 kHz with a 1 kHz high pass
filter (see XX for more information).

Recordings from Hawaii were collected using drifting acoustic recorders
during the 2017 HICEAS (Hawaiian Island Cetacean Ecosystem Assessment
Survey, Yano et al., 2018).~ Drifting recording buoys were deployed at
randomly selected locations within the main Hawaiian Islands. Two types
of drifting recorders were included: (1) Soundtrap ST4300 recorders
(Ocean Acoustics, Auckland, New Zealand) and (2) Wildlife Acoustics SM3M
recorders (Wildlife Acoustics, Maynard, MA). Soundtrap recorders were
sampled at 288 kHz with a duty cycle of 2 min on for every 10 minutes.
SM3M data were sampled at 256 kHz with continuous recordings. Individual
buoys were deployed for 10 to 50 days (see Yano et al., 2018 for more
detailed information).

Recordings from the Eastern North Pacific Ocean off the U.S. west coast
(EPacific) were collected using drifting acoustic recorders during the
2016 PASCAL (Passive Acoustic Survey of Cetacean Abundance Levels)
survey (Keating et al., 2018). Buoys were deployed at a number of
stations in the California Current, including pre-determined locations
(for abundance estimation studies) and in close proximity to seamounts
(Ad Hoc Seamount Experiment). Two types of drifting recorders were
included: (1) Soundtrap ST4300 recorders (Ocean Acoustics, Auckland, New
Zealand) and (2) Wildlife Acoustics SM3M recorders (Wildlife Acoustics,
Maynard, MA). Recordings from pre-determined locations were duty cycled
and recorded 2 minutes out of a variable `off' time; recordings from the
Seamount Experiment were continuous. Soundtrap recorders were sampled at
288 kHz and the SM3M recorders were recorded at 256 sample rate (with
some sampling at 96 kHz for ocean noise measurements). Individual buoys
were deployed for 11-23 days (see Keating et al.~2018 for more detailed
information).

@@

All acoustic recordings were analyzed using PAMGuard software (version
2.00.15c) with a suite of generic click classifiers within the Click
Detector module (see Keating and Barlow, 2013).~ Classifier sets were
saved such that any click may be classified as more than one click type.
The general click classifiers were effectively click detectors, as each
implemented a simple spectral band settings (e.g., 2- 15 kHz, 15 -- 30
kHz, 30 -- 50 kHz, 50 -- 80 kHz, and \textgreater{} 80 kHz). An
additional detector/classifier within the 30 -- 50 kHz peak frequency
range considered the presence of a frequency sweep that is
characteristic of beaked whale pulses (see Keating and Barlow, 2013).

Acoustic detection of beaked whale events were then classified to
species by experienced acoustic researchers using multiple lines of
evidence. If beaked whales were sighted in close proximity to the
recording device, the visual confirmation of species identify was used
to determine species identity. Otherwise, experienced acousticians
followed strict protocol to define species identify based on acoustic
characteristics. When the available characteristics were inconclusive,
the species was considered an unidentified beaked whale; only acoustic
events classified to species were considered for this study. Acoustic
researchers with experience in classifying beaked whale echolocation
clicks examined the bearing-time display for the click detector module
in PAMGuard to assess the species identity of the acoustic event. For
NEFSC data, acoustic detections of beaked whales using towed hydrophone
arrays were further subdivided into click trains based on identification
of consecutive clicks along the same bearing angle in the bearing time
plot (see DeAngelis et al., 2018). Data were saved within the PAMGuard
database and binaries for downstream processing using PAMpal package
(Sakai, 2021) in R programming language (R Core Team 2014).

Standard click calculations were measured from echolocation clicks using
default values in PAMpal. ~For EPacific data, any events \textgreater{}
2 minutes were subdivided into 2-minute events to ensure that all events
were of equal value (because most of our data was duty cycled into
2-minute blocks). Inter-click interval (ICI) was calculated using the
`calculateICI' function in PAMpal; the mode of the ICI was identified
for each beaked whale acoustic event. In addition, for EPacific data,
used the `browseEdinfo' and `matchEnvData' function in PAMpal to
identify appropriate ERDAPP environmental data (cite), including: sea
surface temperature (from jplMURSST41), productivity (from~
~erdMH1pp8day dataset), chlorophyll (from erdMBchla8day dataset), and
seafloor gradient (x,y, and gradient magnitude from
rdSrtm30plusSeafloorGradient).

Multiple trials were conducted to better understand the value (and
potential limitations) of the different suite of measurements. Data
subsets for each trial were exported for BANTER classification using the
export\_banter function in PAMpal. At a minimum, each trial included
standard click calculations for all clicks in a beaked whale acoustic
event (EC); and some trials included the mode of the ICI for each event
(EC\_ICI). For the EPACIFIC data, additional trials that also included
water depth (EC\_ICI\_D) and environmental data (EC\_ICI\_ENV) were run
for each event.

@@

BANTER models was created for each trial using the BANTER package (cite)
in R programming language (R Core Team 2014). Standard click
calculations were used for the BANTER call classifier; event level
measures of ICI and environmental features (for EPacific data) were
applied to the BANTER event classifier. A minimum of two events per
species and complete suite of variables (absence of NAs) were required
for training and testing of the classification model. The BANTER random
forest models contain two parameters, ntree and sampsize, which must be
selected to ensure stability of the model. The sample size (sampsize) is
the number of events considered for each trial of the model, and the
ntree is the number of trees (iterations of the model) run in a given
model. These values were modified to obtain the best results, with an
emphasis on lower sample size and higher number of trees for
computational performance (Rankin and Archer, 2021).

The model output include the confusion matrix, and a number of
visualizations provided by the rfpermute package (cite) in R programming
language (R Core Team 2014).

@@

Four study areas (NAtlantic, SAtlantic, Hawaii, EPacific) were analyzed
for this study; species included: Baird's beaked whales (\emph{Berardius
bairdii}), Blainsville's beaked whales (\emph{Mesoplodon densirostris}),
Cuvier's beaked whales (\emph{Ziphius cavirostris}), Cross Seamount
beaked whale (McDonald et al., 2009), Gervais' beaked whales
(\emph{Mesoplodon europaeus}), Longman's beaked whale (\emph{Indopacetus
pacificus}), Sowerby's beaked whales (\emph{Mesoplodon bidens}), True's
beaked whales (\emph{Mesoplodon mirus}), Stejneger's beaked whale
(\emph{Mesoplodon stejnegeri}), and unidentified beaked whale `BW43'
(Baumann-Pickering et al., 2013). Analyses were conducted separately for
the different study areas due to differences in the data collection
methods that precluded combining of datasets. Specific species
combinations and sample sizes varied by study area.

@@

{[}\emph{Note: Last reprocess had both models with same classification
results (though ICI was `important', it did not really change the
results). I need to reword a few things to accommodate this (and waiting
on confirmation that re: ptoential changes to one function in
PAMpal}){]}

@@

\begin{minipage}[b]{0.50\linewidth}

\begin{longtable}[]{@{}lllll@{}}
\toprule()
& Chinstrap & Gentoo & Adelie & \textbf{\% Correct} \\
\midrule()
\endhead
Chinstrap & 70 & 0 & 0 & 100.0 \\
Gentoo & 0 & 5 & 0 & 100.0 \\
Adelie & 0 & 0 & 65 & 100.0 \\
\textbf{Overall} & - & - & - & 100.0 \\
\bottomrule()
\end{longtable}

\end{minipage}%
%
\begin{minipage}[b]{0.50\linewidth}

\begin{longtable}[]{@{}lll@{}}
\toprule()
LCI 0.95 & UCI 0.95 & Prior \\
\midrule()
\endhead
94.9 & 100.0 & 50.0 \\
47.8 & 100.0 & 3.6 \\
94.5 & 100.0 & 46.4 \\
97.4 & 100.0 & 46.7 \\
\bottomrule()
\end{longtable}

\end{minipage}%
\newline
\begin{minipage}[b]{0.33\linewidth}

\begin{enumerate}
\def\labelenumi{\alph{enumi})}
\tightlist
\item
  \includegraphics{Figures/myplot1.jpg}
\end{enumerate}

\end{minipage}%
%
\begin{minipage}[b]{0.33\linewidth}

\begin{enumerate}
\def\labelenumi{\alph{enumi})}
\setcounter{enumi}{1}
\tightlist
\item
  \includegraphics{Figures/myplot2.jpg}
\end{enumerate}

\end{minipage}%
%
\begin{minipage}[b]{0.33\linewidth}

\begin{enumerate}
\def\labelenumi{\alph{enumi})}
\setcounter{enumi}{2}
\tightlist
\item
  \includegraphics{Figures/myplot3.jpg}
\end{enumerate}

\end{minipage}%
\newline
\begin{minipage}[b]{0.33\linewidth}

@@

::: kable(hawaii/hawaii\_confuseMatrix) :::

\begin{enumerate}
\def\labelenumi{\alph{enumi})}
\tightlist
\item
  \href{hawaii/hawaii_proximity.png}{Fig.1}
\end{enumerate}

\begin{enumerate}
\def\labelenumi{\alph{enumi})}
\setcounter{enumi}{1}
\tightlist
\item
  \href{hawaii/hawaii_importance.png}{Fig.2}
\end{enumerate}

\begin{enumerate}
\def\labelenumi{\alph{enumi})}
\setcounter{enumi}{2}
\tightlist
\item
  \href{hawaii/hawaii_votes.png}{Fig.3}
\end{enumerate}

@@

The EPacific dataset consisted of 25 drifting recording buoys (20
stations plus 5 seamount experiment drifts) deployed off the west coast
of the United States (Keating et al.~2018); species included Baird's
beaked whale (n=91), Unidentified Beaked Whale `BW43' (n=11),
Stejneger's beaked whale (n=41) and Cuvier's beaked whale (n=1,332).
BANTER classification trials included (1) echolocation clicks (EC), (2)
echolocation clicks and ICI (EC\_ICI), (3) echolocation clicks, ICI, and
water depth (EC\_ICI\_D), and (4) echolocation clicks, ICI, water depth,
sea surface salinity, productivity, and chlorophyll (EC\_ICI\_ENV).

@@

Previous application of BANTER to dolphin species in the California
Current found that large sample sizes could result in strong
classification of species where experienced acousticians are unable to
differentiate species (long-beaked and short-beaked common dolphins,
Rankin et al., 2017). This suggests that large increases in sample size
may improve classification results for Cuvier's beaked whales.

@@

Gervais' and True's beaked whales have similar click characteristics
that make them difficult to differentiate, and require considerable
expertise to classify manually, using a variety of available visuals and
metrics (DeAngelis et al., 2018). Our results suggest that BANTER may
serve as an efficient and effective means of classifying and
differentiating these two species. Despite modest sample sizes, Gervais'
and True's beaked whales showed high classification scores (95.6\% and
100\%, respectively). Unfortunately, we were unable to combine the N and
S Atlantic datasets (which would increase sample sizes of True's beaked
whales) due to differences in hydrophone characteristics that result in
differences in call measures. Calibration of signals (to make them
comparable) may allow for combining datasets to improve sample sizes.

For the SAtlantic models, ICI was found to be an important predictor for
many species {[}\emph{talk about the 1 change in sowerby's when the ICI
was added- do you think it was because of an error in ICI?}{]} . For
Hawaii and the EPacific, however, we found that adding ICI into the
BANTER classifier did not improve final classification results. The
differences here are likely due to different analysis methods for
identifying events. Beaked whale events were defined by encounter time
for both the Hawaii and EPacific data, which were collected using
drifting acoustic recorders where animals could not be localized. The
Atlantic data were collected using a towed hydrophone array with
localization capabilities. Each event was defined as an individual click
train, so ICI for these events more likely represented true ICI, and was
found to be more informative. Baumann-Pickering et al.~(2013) showed an
overlap in the ICI between Cuvier's and Blainsville's beaked whales, but
a strong difference in ICI between these species and BWC. Likewise,
Baumann-Pickering (2013) found a difference in ICI between Baird's
beaked whales and Cuvier's beaked whales. Subdividing click trains in
PAMGuard is time consuming, but alternative options include
consideration of the PAMGuard click train detector module in BANTER, or
by developing a similar function in R to apply to events. Integration of
ICI from individual click trains may serve to improve classification
results for some species in Hawaii and EPacific.

Gervais were eliminated from the NAtlantic BANTER model due to low
sample size (n=2); results were extremely poor when they were included
in the model. Analysis notes indicate uncertainty in manual species
classification on one of the two events in this dataset. While small
sample size can at times provide good classification scores, as we found
in Sowerby's beaked whale in this SAtlantic dataset, they can be
devastating if there are inaccuracies in the training data. These two
examples of small sample sizes highlights a conflict in preferred
protocol. Ideally, (1) all \uline{species} would be included in a
classification model, so as to better represent the local species
diversity, (2) all \uline{events} would ideally be included in the
classification model, so as to better represent the variability found in
the area, and (3) only confident `ground truth' classifications would be
considered in the training model. Unfortunately, in species where
identity must be determined based on call characteristics (rather than
visual confirmation of species identity in the field), it can sometimes
be difficult to confidently determine species identity when calls are
highly variable and do not include at least one call that provides high
confidence of species identity. In the case of beaked whales, we
recommend that training data include a reasonable level of confidence
for inclusion. When sample sizes are low, such as in these cases, it may
be wise to consider the strength of the classifications when including
them for model training (as we have done here). An alternative is to
require agreement from multiple analysts, as provided by the EPacific
datasets. This concern about accurate labeling of training data is
further complicated by the potential existence of species that have not
yet been identified. For example, Barlow et al.~(2021b) recently
discovered what appears to be a new species of beaked whale off Baja
California, Mexico; this putative new species may have been detected but
misclassified in other datasets (including the EPacific data presented
here).

@@

Here are two sample references: Feynman and Vernon Jr. (1963) Dirac
(1953).

>>>>

\title{Open-Source machine learning BANTER acoustic classification of a
cryptic echolocating species}
\author{Shannon Rankin \and Taiki Sakai \and Frederick Archer \and Jay
Barlow \and Danielle Cholewiak \and Annamaria DeAngelis \and Jennifer L.
K. McCullough \and Erin Oleson \and Melissa Soldevilla}
\date{2/6/23}

\begin{document}
\maketitle
\begin{abstract}
Passive acoustic monitoring is increasingly used for assessing
populations of marine mammals; however, analysis of large datasets is
limited by our ability to easily classify sounds detected.
Classification of beaked whale acoustic events, in particular, require
evaluation of multiple lines of evidence by expert analysts. Here we
present a highly automated approach to acoustic detection and
classification using supervised machine learning and open source
software methods. Data from four large scale surveys of beaked whales
(North Atlantic, South Atlantic, Hawaii, and Eastern Pacific) were
analyzed PAMGuard (acoustic detection), PAMpal (acoustic analysis) and
BANTER (hierarchical random forest classifier). Overall correct
classification results ranged from 86\% for the South Atlantic data to
99\% for the North Atlantic. Results for many species could likely be
improved with increased sample sizes and consideration of alternative
automated detectors. These methods provide a highly automated approach
to acoustic detection and classification using open source methods that
can be readily adopted for species and geographic regions.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, sharp corners, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt]}{\end{tcolorbox}}\fi

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(tidyverse)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
-- Attaching packages --------------------------------------- tidyverse 1.3.2 --
v ggplot2 3.4.0      v purrr   1.0.1 
v tibble  3.1.8      v dplyr   1.0.10
v tidyr   1.2.1      v stringr 1.5.0 
v readr   2.1.3      v forcats 1.0.0 
-- Conflicts ------------------------------------------ tidyverse_conflicts() --
x dplyr::filter() masks stats::filter()
x dplyr::lag()    masks stats::lag()
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(knitr)}
\FunctionTok{library}\NormalTok{(here)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
here() starts at C:/Users/shannon.rankin/Documents/GitHub/BANTER_BeakedWhales
\end{verbatim}

@@

Passive acoustic monitoring (PAM) from has proven to be a valuable tool
for studying populations of marine mammals ((Parijs et al. 2009));
however, the value of these studies depends on our ability to identify
the sources of the sounds we are monitoring. Historically, experienced
acousticians manually identify stereotyped sounds that could be reliably
attributed to a given species, based on their spectral or temporal
characteristics ((Baumann-Pickering et al. 2013a); (Bittle and Duncan
2013); (Rankin and Barlow 2005); (Soldevilla et al. 2008)). The dramatic
increase in recordings make it impossible for experienced acousticians
to manually annotate all data, and there has been an increasing need for
the development of automated classification routines that can provide
accurate determinations of the species responsible for the sounds.

Beaked whales are deep diving marine mammals found in offshore waters;
their long dive intervals and cryptic surfacing behavior make them
difficult to study using typical shipboard visual observation methods
((MacLeod 2018)). However, beaked whales make stereotyped echolocation
clicks that have temporal and spatial characteristics that can be used
to differentiate species ((Baumann-Pickering et al. 2013b)). Detection
of beaked whale signals in large datasets is greatly expanding our
understanding of their population structure and the potential impact of
human activities on these species ((Barlow et al. 2022);
(Baumann-Pickering et al. 2014); (Simonis et al. 2020)). Unfortunately,
manual classification of beaked whale echolocation clicks requires
analysis by trained observers using a number of visual aids to examine
the call characteristics. Development of automated classification
routines, if accurate, serve to improve the efficiency and decrease the
cost of analyzing large datasets.

BANTER is a supervised machine learning acoustic event classifier
originally developed for classifying dolphin schools from acoustic
recordings ((Rankin et al. 2017a)). This hierarchical random forest
event classifier relies on an initial call classifier (in the dolphin
case study, this included one each for whistles, echolocation clicks,
and burst pulses) followed by an event classifier. The event classifier
considers the probability of assignment to species for each call in the
call classifier, along with any event-level characteristics (such as
call rate). BANTER is very flexible and can accommodate any number of
measures from any number of detectors. While BANTER relies on multiple
call types, relatively minor changes to a call detector could yield
different detector results ((Rankin et al. 2017b)). In this case,
settings for a whistle \& moan detector were modified to improve
performance at detecting burst pulses. This approach of applying
multiple call detectors (of the same type but with different settings)
suggests that perhaps BANTER could be used on species that only produce
one call type (or where only one call type was analyzed), as long as the
settings of the automated call detector were modified such that the
overall number of detections was different. This approach was
successfully applied to classified echolocation clicks for narwhal and
belugas ((Zahn et al. 2021)). For species who primarily (or exclusively)
produce echolocation clicks, BANTER may serve as an option for automated
machine learning classification.

@@

Recordings from the western North Atlantic Ocean (U.S. east coast) were
collected using towed hydrophone arrays during the 2016 AMAPPS II
(Atlantic Marine Assessment Program for Protected Species) survey
(cite). AMAPPS II was subdivided into the northern (NAtlantic) and
southern (SAtlantic) survey area. The NAtlantic study area ranged from
Massachusetts south to New Jersey (HB1603, {[}CITE{]}) and recordings
used in this analysis consisted of a single hydrophone (HTI-96-min, High
Tech Inc., Long Beach, MS) recorded at 192 kHz sample rate with a 1 kHz
high pass filter (National Instruments USB-6356 A/D card, see (DeAngelis
et al. 2018a) for more information). The SAtlantic area ranged from
Delaware south to central Florida (GU1605, {[}CITE{]}) and recordings
used in this analysis consisted of a single hydrophone (Reson
{[}MODEL{]}, Teledyne Marine, Slangerup, Denmark) recorded 500 kHz
sample rate (soundcard) and decimated to 192 kHz with a 1 kHz high pass
filter (see {[}CITE{]} for more information).

Recordings from Hawaii were collected using drifting acoustic recorders
during the 2017 HICEAS (Hawaiian Island Cetacean Ecosystem Assessment
Survey, (Yano et al. 2018)).~ Drifting recording buoys were deployed at
randomly selected locations within the main Hawaiian Islands. Two types
of drifting recorders were included: (1) Soundtrap ST4300 recorders
(Ocean Acoustics, Auckland, New Zealand) and (2) Wildlife Acoustics SM3M
recorders (Wildlife Acoustics, Maynard, MA). Soundtrap recorders were
sampled at 288 kHz with a duty cycle of 2 min on for every 10 minutes.
SM3M data were sampled at 256 kHz with continuous recordings. Individual
buoys were deployed for 10 to 50 days (see (Yano et al. 2018)for more
detailed information).

Recordings from the Eastern North Pacific Ocean off the U.S. west coast
(EPacific) were collected using drifting acoustic recorders during the
2016 PASCAL (Passive Acoustic Survey of Cetacean Abundance Levels)
survey ((J. Keating et al. 2018a)). Buoys were deployed at a number of
stations in the California Current, including pre-determined locations
(for abundance estimation studies) and in close proximity to seamounts
(Ad Hoc Seamount Experiment). Two types of drifting recorders were
included: (1) Soundtrap ST4300 recorders (Ocean Acoustics, Auckland, New
Zealand) and (2) Wildlife Acoustics SM3M recorders (Wildlife Acoustics,
Maynard, MA). Recordings from pre-determined locations were duty cycled
and recorded 2 minutes out of a variable `off' time; recordings from the
Seamount Experiment were continuous. Soundtrap recorders were sampled at
288 kHz and the SM3M recorders were recorded at 256 sample rate (with
some sampling at 96 kHz for ocean noise measurements). Individual buoys
were deployed for 11-23 days (see (J. Keating et al. 2018b) for more
detailed information).

@@

All acoustic recordings were analyzed using PAMGuard software (version
2.00.15c) with a suite of generic click classifiers within the Click
Detector module (see (J. L. Keating and Barlow 2013a)).~ Classifier sets
were saved such that any click may be classified as more than one click
type. The general click classifiers were effectively click detectors, as
each implemented a simple spectral band settings (e.g., 2- 15 kHz, 15 --
30 kHz, 30 -- 50 kHz, 50 -- 80 kHz, and \textgreater{} 80 kHz). An
additional detector/classifier within the 30 -- 50 kHz peak frequency
range considered the presence of a frequency sweep that is
characteristic of beaked whale pulses (see (J. L. Keating and Barlow
2013b)).

Acoustic detection of beaked whale events were then classified to
species by experienced acoustic researchers using multiple lines of
evidence. If beaked whales were sighted in close proximity to the
recording device, the visual confirmation of species identify was used
to determine species identity. Otherwise, experienced acousticians
followed strict protocol to define species identify based on acoustic
characteristics. When the available characteristics were inconclusive,
the species was considered an unidentified beaked whale; only acoustic
events classified to species were considered for this study. Acoustic
researchers with experience in classifying beaked whale echolocation
clicks examined the bearing-time display for the click detector module
in PAMGuard to assess the species identity of the acoustic event. For
NEFSC data, acoustic detections of beaked whales using towed hydrophone
arrays were further subdivided into click trains based on identification
of consecutive clicks along the same bearing angle in the bearing time
plot (see (DeAngelis et al. 2018b)). Data were saved within the PAMGuard
database and binaries for downstream processing using PAMpal package
((Sakai 2021)) in R programming language ((R Core Team 2022)).

Standard click calculations were measured from echolocation clicks using
default values in PAMpal. For EPacific data, any events \textgreater{} 2
minutes were subdivided into 2-minute events to ensure that all events
were of equal value (because most of our data was duty cycled into
2-minute blocks). Inter-click interval (ICI) was calculated using the
`calculateICI' function in PAMpal; the mode of the ICI was identified
for each beaked whale acoustic event.

Multiple trials were conducted to better understand the value (and
potential limitations) of the different suite of measurements. Data
subsets for each trial were exported for BANTER classification using the
export\_banter function in PAMpal. At a minimum, each trial included
standard click calculations for all clicks in a beaked whale acoustic
event (EC); and some trials included the mode of the ICI for each event
(EC\_ICI).

@@

BANTER models was created for each trial using the BANTER package in R
programming language ((Archer 2022a)). Standard click calculations were
used for the BANTER call classifier; event level measures of ICI were
applied to the BANTER event classifier. A minimum of two events per
species and complete suite of variables (absence of NAs) were required
for training and testing of the classification model. The BANTER random
forest models contain two parameters, ntree and sampsize, which must be
selected to ensure stability of the model. The sample size (sampsize) is
the number of events considered for each trial of the model, and the
ntree is the number of trees (iterations of the model) run in a given
model. These values were modified to obtain the best results, with an
emphasis on lower sample size and higher number of trees for
computational performance ((Rankin and Archer 2021)).

The model output include the confusion matrix, and a number of
visualizations provided by the rfpermute package in R programming
language ((Archer 2022b)).

@@

Four study areas (NAtlantic, SAtlantic, Hawaii, EPacific) were analyzed
for this study; species included: Baird's beaked whales (\emph{Berardius
bairdii}), Blainsville's beaked whales (\emph{Mesoplodon densirostris}),
Cuvier's beaked whales (\emph{Ziphius cavirostris}), Cross Seamount
beaked whale (McDonald et al., 2009), Gervais' beaked whales
(\emph{Mesoplodon europaeus}), Longman's beaked whale (\emph{Indopacetus
pacificus}), Sowerby's beaked whales (\emph{Mesoplodon bidens}), True's
beaked whales (\emph{Mesoplodon mirus}), Stejneger's beaked whale
(\emph{Mesoplodon stejnegeri}), and unidentified beaked whale `BW43'
((Baumann-Pickering et al. 2013a)). Analyses were conducted separately
for the different study areas due to differences in the data collection
methods that precluded combining of datasets. Specific species
combinations and sample sizes varied by study area.

@@

@@

@@

\begin{minipage}[t]{\linewidth}

\begin{enumerate}
\def\labelenumi{\alph{enumi})}
\item
  \begin{figure}

  {\centering \includegraphics{../hawaii/banter_model_hawaii_confuseMatrix.png}

  \caption{\label{fig-a}Fig.1}

  \end{figure}
\item
  \includegraphics{../hawaii/banter_model_hawaii_proximity.png} :::
\item
  \begin{figure}

  {\centering \includegraphics{../hawaii/banter_model_hawaii_importance.png}

  \caption{Fig.1}

  \end{figure}
\item
  \begin{figure}

  {\centering \includegraphics{../hawaii/banter_model_hawaii_votes.png}

  \caption{Fig.1}

  \end{figure}
\end{enumerate}

@@

The EPacific dataset consisted of 25 drifting recording buoys (20
stations plus 5 seamount experiment drifts) deployed off the west coast
of the United States ((J. Keating et al. 2018a)); species included
Baird's beaked whale (n=91), Unidentified Beaked Whale `BW43' (n=11),
Stejneger's beaked whale (n=41) and Cuvier's beaked whale (n=1,332).
BANTER classification trials included (1) echolocation clicks (EC) and
(2) echolocation clicks and ICI (EC\_ICI).

@@

Previous application of BANTER to dolphin species in the California
Current found that large sample sizes could result in strong
classification of species where experienced acousticians are unable to
differentiate species (long-beaked and short-beaked common dolphins,
(Rankin et al. 2017c)). This suggests that large increases in sample
size may improve classification results for Cuvier's beaked whales.

@@

Gervais' and True's beaked whales have similar click characteristics
that make them difficult to differentiate, and require considerable
expertise to classify manually, using a variety of available visuals and
metrics ((DeAngelis et al. 2018c)). Our results suggest that BANTER may
serve as an efficient and effective means of classifying and
differentiating these two species. Despite modest sample sizes, Gervais'
and True's beaked whales showed high classification scores (95.6\% and
100\%, respectively). Unfortunately, we were unable to combine the N and
S Atlantic datasets (which would increase sample sizes of True's beaked
whales) due to differences in hydrophone characteristics that result in
differences in call measures. Calibration of signals (to make them
comparable) may allow for combining datasets to improve sample sizes.

For the SAtlantic models, ICI was found to be an important predictor for
many species {[}\emph{talk about the 1 change in sowerby's when the ICI
was added- do you think it was because of an error in ICI?}{]} . For
Hawaii and the EPacific, however, we found that adding ICI into the
BANTER classifier did not improve final classification results. The
differences here are likely due to different analysis methods for
identifying events. Beaked whale events were defined by encounter time
for both the Hawaii and EPacific data, which were collected using
drifting acoustic recorders where animals could not be localized. The
Atlantic data were collected using a towed hydrophone array with
localization capabilities. Each event was defined as an individual click
train, so ICI for these events more likely represented true ICI, and was
found to be more informative. (Baumann-Pickering et al. 2013a) showed an
overlap in the ICI between Cuvier's and Blainsville's beaked whales, but
a strong difference in ICI between these species and BWC. Likewise,
(Baumann-Pickering et al. 2013a) found a difference in ICI between
Baird's beaked whales and Cuvier's beaked whales. Subdividing click
trains in PAMGuard is time consuming, but alternative options include
consideration of the PAMGuard click train detector module in BANTER, or
by developing a similar function in R to apply to events. Integration of
ICI from individual click trains may serve to improve classification
results for some species in Hawaii and EPacific.

Gervais were eliminated from the NAtlantic BANTER model due to low
sample size (n=2); results were extremely poor when they were included
in the model. Analysis notes indicate uncertainty in manual species
classification on one of the two events in this dataset. While small
sample size can at times provide good classification scores, as we found
in Sowerby's beaked whale in this SAtlantic dataset, they can be
devastating if there are inaccuracies in the training data. These two
examples of small sample sizes highlights a conflict in preferred
protocol. Ideally, (1) all \uline{species} would be included in a
classification model, so as to better represent the local species
diversity, (2) all \uline{events} would ideally be included in the
classification model, so as to better represent the variability found in
the area, and (3) only confident `ground truth' classifications would be
considered in the training model. Unfortunately, in species where
identity must be determined based on call characteristics (rather than
visual confirmation of species identity in the field), it can sometimes
be difficult to confidently determine species identity when calls are
highly variable and do not include at least one call that provides high
confidence of species identity. In the case of beaked whales, we
recommend that training data include a reasonable level of confidence
for inclusion. When sample sizes are low, such as in these cases, it may
be wise to consider the strength of the classifications when including
them for model training (as we have done here). An alternative is to
require agreement from multiple analysts, as provided by the EPacific
datasets. This concern about accurate labeling of training data is
further complicated by the potential existence of species that have not
yet been identified. For example, (cite jays new paper) recently
discovered what appears to be a new species of beaked whale off Baja
California, Mexico; this putative new species may have been detected but
misclassified in other datasets (including the EPacific data presented
here).

@@

Here are two sample references: (\textbf{Feynman1963118?})
(\textbf{Dirac1953888?}).
